---
title: "Final Project"
author:
- Jessie Bierschenk
- Aidan Gildea
- Julia Mitchell
- Naomi Rubin
- Gaurav Sirdeshmukh
output:
  pdf_document: default
  html_document: default
---

```{r setup, include = FALSE}
library(dplyr)
library(tidyr)
library(magrittr)
library(rvest)
library(readr)
library(tidyverse)
library(purrr)
library(XML)
library(janitor)
library(stringr)
library(ggplot2)
library(kableExtra)
```

## Introduction

This year, Duke witnessed a historic basketball season, with it being Coach K's last season and playing so well throughout the season and March Madness. Therefore, we decided to make our final project all about Duke men's basketball statistics. Our project scrapes Duke men's basketball data, about both the team and individual players, from the goduke.com website. We then created a Shiny app to present this data and allow users to look up individual player stats and access different summary statistic reports. The app incorporates visualizations relevant to team and individual data. Our goal is to create an app that provides the user with informative and accessible insights into team and player performance, while also being reproducible for future seasons.

## Methods / Implementation

The first half of our project is dedicated to scraping and cleaning the Duke
Men's basketball data from the 2021-2022 season. We scraped this data from
`https://goduke.com/sports/mens-basketball/stats`. For our project, we wanted to
scrape that data on the team statistics and overall individual statistics. We 
used selector gadget and inspected the pages to determine the best way to scrape 
the data. We were able to find the xpath by inspecting the pages and hovering 
over the element we wanted to scrape. We used xpaths that returned all of the
data in html format for the "Team" and "Individual" pages. After scraping the
data we had to clean it.

We split the team and individual statistics into two separate data frames. For 
the team statistics data frame, we had to clean the data in multiple ways. First,
we added a variable "type" to indicate which category, (i.e. "Scoring", 
"Shooting","Rebounding", etc.) the statistic falls under. We populated this
variable for every statistics. We also renamed poorly labelled columns
and cleaned the output of the rows. 

For the individual statistics data frame, we also had to clean the data. First,
we renamed the columns.The columns specify the statistic, unit, and the category 
the statistic falls into ("minutes", "scoring", and "rebounds") if it is a repeated 
column name (i.e. "AVG"). We also fixed the rows so that the "Player" column
returns the row in a better format. Lastly,the html contains "conference", 
"scoring", and "average" statistics in addition to "overall" but we only 
included the "overall" data. We filtered the data frame for the overall data
and removed all rows that were not in that data set.

Then we developed our Shiny app. First, the user can interact with a drop down menu on the side panel and choose to view team stats or choose to view individual player stats. If the user chooses to view individual player stats, they then can interact with another drop down menu and choose from the roster which player's stats they would like to view. Upon pressing the action button, the main panel will produce the output corresponding to the appropriate selection. In the main panel, users can toggle between two panels, one for the overall stat line (in table format) or a summary visualization (in graph format). The content of the main panel will update dynamically when switching between the two tabs; however, if changing the input for the team/individual drop down, the user must press search again. The overall stat line is a table of statistics, either about the team's performance or an individual player's performance depending on the user's selection in the side panel. The summary visualization tab features a unique plot for team performance and a dynamic plot for individual performance. The graph for team performance allows users to compare the team's per game average statistics (ie. points per game, rebounds per game). The graph for individual performance allows users to compare the players' FG%, FT%, and 3PT%, with the player of interest's data points highlighted. Altogether, the app is an intuitive and east-to-use tool for users who want to explore the Duke Men's Basketball Team's performance during the 2021-2022 season. 

## Discussion & Conclusions

After scraping the data and creating a Shiny app, we feel that we have presented Duke men's basketball statistics in an informative, interactive, and user-friendly way. The app allows users to filter the data easily and look at a variety of different summary statistics. Although we are happy with the way that our app turned out, we feel that there is much we could expand upon in the future.

In the future, we would be interested in expanding upon our project by scraping more data and adding more elements to our Shiny app. This could look like scraping data from previous years, scraping data about other teams, adding more visualizations to our Shiny app, and generally expanding the scope of our app.

One of our main objectives was to ensure that our code is reproducible and generalizable. Every ACC school uses the same website for their athletics pages, and as such, team and individual data is formatted essentially identically. We could expand this app from just Duke basketball, to all of the ACC - and do so without having to worry about data frames being structured differently. This could open the door to many new visualizations and for comparisons between teams and our app would become more of an analytical tool than it currently is. An alternative to this would be to implement a fileInput() where the user can scrape and import data themselves. However, doing this scraping and cleaning ourselves and updating the app’s functionality to allow for team selection would benefit the user experience.

### Data Wrangling: Team Statistics

```{r}
url <- read_html("https://goduke.com/sports/mens-basketball/stats")
# xpath found using inspect and hovering over element
team_stats <- url %>% html_nodes(xpath='//*[@id="team"]') %>% html_table()
# Create "Type" variable to indicate which category of statistics
type <- c("Scoring", "Shooting","Rebounding","Assists","Turnovers","Steals",
          "Blocks", "Attendance")
team_stats <- as.data.frame(team_stats) %>% mutate("Type" = NA)
# Extract the name of the statistics
team_stats$Statistic <- sub("\\r.*", "", team_stats$Statistic)
# Rename the "Opponents" column
names(team_stats)[names(team_stats) == "Opponents..............................OPP"] <- "Opponent"
# Populate the "type" variable
tally <- 0
for (i in 1:nrow(team_stats)){
  match <- match(team_stats$Duke[i], type[tally+1])
  if (is.na(match) == FALSE) {tally = tally + 1}
  team_stats$Type[i] = type[tally]
  if (team_stats$Statistic[i] == "Per Game" |
      team_stats$Statistic[i] == "Total"
      | team_stats$Statistic[i] == "Margin") {
    team_stats$Statistic[i] <-
      paste0(team_stats$Type[i], " ", team_stats$Statistic[i])
  }
}
# Move the type variable to an earlier column
team_stats <- team_stats[!team_stats$Duke %in% type, ] %>% relocate(Type, .after = Statistic)
knitr::kable(head(team_stats))
```


### Data Wrangling: Team Statistics

```{r}
# xpath found using inspect and hovering over element
ind_stats <- url %>% html_nodes(xpath='//*[@id="individual"]') %>% html_table()
# Save as data frame and remove irrelevant variables
ind_stats <- as.data.frame(ind_stats) %>% select(-Bio.Link, -Var.28)
type2 <- c("Minutes", "Scoring", "Rebounds")
# Fix the names of the columns to include variable and units
names(ind_stats) <- paste(names(ind_stats)," ",as.character(unlist(ind_stats[1,])))
# Extract the proper column name from the header 
# Indicate the type of stat when needed

for (i in 1:ncol(ind_stats)){
  results <- lapply(type2, function(x)str_extract(names(ind_stats)[i], x))
  if (all(is.na(results))==TRUE) {
    names(ind_stats)[i] <- sub("^\\S+\\s+", "",names(ind_stats)[i])
    }
}

# Extract the player name from the "Player" column
ind_stats$Player<- unlist(lapply(ind_stats$Player, function(x)str_extract(x, "[^\r]+")))
# Remove the first row with that indicates units
ind_stats <- ind_stats[-1,]
# Rename the player number column as "Number"
names(ind_stats)[names(ind_stats) == "X."] <- "Number"
# Filter for the "Overall" individual stats
remove <- 0
for (i in seq(nrow(ind_stats))){
  if (ind_stats$Player[i] == "Total"){
    remove <- i
    break
  }
}
# Remove rows that are not the overall individual stats
ind_stats <- ind_stats[-c(i:nrow(ind_stats)), ]
team_table <- ind_stats[,1:6]
knitr::kable(head(team_table))

# Use for shiny app
player_names <- ind_stats %>%
  select(Player) %>%
  unlist(use.names = FALSE) %>%
  unique()
```

### Shiny App

```{r shinyApp}
library(shiny)
library(purrr)
library(htmltools)

# create a shiny app
shinyApp(
  # create user interface (UI)
  ui = fluidPage(
    # app title and subtitle
    titlePanel(HTML("Duke Men's Basketball 21-22", "<br>", "<font size=5><i>2021-2022 Season</i></font>")),
    sidebarLayout(
      # create sidebar panel
      sidebarPanel(
        h4("Statistic Type:"),
        # drop down selection for team or individual statistics
        selectInput("stats", "Choose Statistics Type", c("Team", "Individual")),
        # panel conditional on individual being selected
        conditionalPanel(
          # individual selected as condition
          condition = "input.stats== 'Individual'",
          h4("Select Player"),
          # select from list of players
          selectInput("player", "Player:", choices = ind_stats$Player)
        ),
        # action button initiating output creation
        actionButton("submit", "Search")
      ),
      # create main panel
      mainPanel(
        # create tabs
        tabsetPanel(id = "tabSetPane", 
                    # tab panel for overall season statistic table
                    tabPanel("Statline", value = "stat",
                             h3("Season Statline"),
                            # table output
                             uiOutput("table")
                    ), 
                    # tab panel for summary statistic visualization
                    tabPanel("Summary Visualization", value = "summary",
                             h3("Season Summary"),
                             # graph output
                             plotOutput("graph")
                    )
        )
      ))),
  # create working server
  server = function(input, output, session) {
    
    # actions dependent on search button being pressed
    observeEvent(input$submit, { #depending on statistic selection 
      
      # use team statistics if team is selected
      if (input$stats == "Team" ) {
        input_stats <- team_stats
      }
      # use individual statistics if individual is selected
      else {
        input_stats <- ind_stats %>%
          # filter for selected plater
          filter(Player == input$player) %>% t() %>% as.data.frame() %>%
          mutate(Statistic = names(ind_stats)) %>% relocate(Statistic, before = V1)
        names(input_stats) <- NULL
      }
      # output table of appropriate statistics
      output$table <- renderTable(input_stats, na = "NA")
    })
    
    # actions dependent on search button being pressed
    observeEvent(input$submit, {
      # create summary visual for team performance if team is selected
      if (input$stats == "Team" ) {
        # retrieve per game statistics (exluding attendance)
        stats_pg <- team_stats %>% 
          filter(grepl('Per Game', Statistic), !grepl('Attendance', Statistic)) %>%
          select(Statistic, Duke, Type)
        # convert variable values to type numeric
        stats_pg$Duke <- as.numeric(stats_pg$Duke)
        
        # use ggplot to create visualization
        plot <- ggplot(stats_pg, aes(x=Statistic, y=Duke)) + 
          geom_point(stat = "identity", size=1, color = "blue") + 
          # add labels to stats
          geom_text(data=stats_pg,
                    aes(label=Duke), nudge_x=0, nudge_y=5, 
                    size = 3.5, show.legend = FALSE, color = "blue") + 
          # create line connecting values to axis (~ similar to bar plot)
          geom_segment(aes(x=Statistic, 
                           xend=Statistic, 
                           y=0, 
                           yend=Duke), size = 10, lineend = "round",
                       color = "blue") +
          # label the graph appropriately
          labs(title="Duke Performance Per Game", 
               subtitle = substitute(paste(italic("2021-2022 Season"))),
               x = "Statistic (per game)",
               y = "Value") +
          # flip coordinates
          coord_flip()
      }
      # create summary visual for individual performance if individual is selected
      if (input$stats == "Individual") {
        # created highlight variable, where selected player is assigned "highlight"
        ind_stats$highlight <- ifelse(ind_stats$Player == input$player, "highlight", "normal")
        
        # retrieve selected players FG%, FT%, and 3PT%
        player <- ind_stats %>%
          filter(Player == input$player) %>%
          select(`FG%`, `FT%`, `3PT%`)
        
        # convert values to type numeric without removing column names
        play_stats <- data.frame(lapply(player, function(x) as.numeric(as.character(x))),
                   check.names=F)

        # use ggplot to create visualization
        plot <- ggplot(data = ind_stats,
                       # create  point for FG% in circle shape
                       mapping = aes(x= reorder(Player, as.numeric(`FG%`), mean),
                                     y=as.numeric(`FG%`),
                                     color = highlight), shape = 20) +
          scale_color_manual("Player", values =  c("highlight" = "blue", "normal" = "grey50"))+
          # adjust point size
          geom_point(size = 2) +
          # flip coordinates
          coord_flip() +
          # create point for FT% in triangle shape
          geom_point(data = ind_stats,
                     mapping = aes(x=Player,
                                   y=as.numeric(`FT%`)), shape = 15, size = 2) +
          # create point for 3PT% in square shape
          geom_point(data = ind_stats,
                     mapping = aes(x=Player,
                                   y=as.numeric(`3PT%`)), shape = 17, size = 2) + 
          # remove legend
          theme(legend.position="none") +
          # give axis appropriate titles
          # title and caption are dynamic to represent selected player
          # caption updates to reflect which stat the player performed best in
          labs(title = paste0(sub(".*, ", "", input$player), " ", sub("\\,.*", "", input$player), "'s Season Statistics"), x="Player",  y="Percentage",
               subtitle = "Circle: FG%, Triangle: FT%, Square: 3PT%",
               caption = paste0("Key Stat: ", sub("\\,.*", "", input$player), " performed best in ", names(play_stats)[which(play_stats == max(play_stats), arr.ind=T)[, "col"]], " with a percentage of ", max(play_stats), ".")
               )
      }
      # output summary visualization
      output$graph <- renderPlot(plot,  width = "auto",
                                 height = "auto")
    })
  }
)
```
